Fused regression for multi-source gene regulatory network inference

### 2016-01-15 paper1.tex
hi

@@

\noindent The matrix $\beta$ can be solved using linear regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring multiple networks.

@@

@@

In order to avoid this difficulty, we use depth-first search to identify linked columns of each TF expression matrix $Y^{S_j}$, then forming design and response matrices through vectorization. We can then incorporate fusion constraints as in the case of single-source single response-variable fused regression. In most cases, we have found the direct solve using augmented matrices to be adequate. This is because the structure of orthology tends to be sparse, so only a small number of genes must be solved at once. However, in the general case of $c_i$ response variables from each of $i \in \text{species}$ data sources the design matrix to be solved is of size $\displaystyle \sum_{c_i} rows(X^{S_i}) \times \displaystyle \sum_{c_i} cols(X^{S_i})$. Although the sparse structure of the augmented design matrix makes this somewhat more tractable, it is still impractical for large constraint chains.

@@

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthology information is faulty, to mimic orthologs which are not functionally analogous  (Figure 4). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``incorrect'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``incorrect'' mappings. We verified that adaptive-fusion unfused mostly ``incorrect'' interactions, while leaving truly analogous interactions fused (Figure 4a. Note that while pairs of interactions between true orthologs are sampled randomly so that the difference in interactions tends to be small, and pairs of interactions between incorrect orthologs are sampled randomly so that the difference in interactions tends to be large, there is an element of randomness in the sampling. This results in a small fraction of interactions between incorrect othologs being similar, and similarly, a small fraction of interactions between correct orthologs being different. Consequently, some interactions between incorrect orthologs remain fused and some interactions between correct orthologs are unfused.

We compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information (Figure 4b). Because fused L2 heavily penalizes large differences between weights which are predicted to be similar, it is able to retrieve a more accurate network for those interactions with true fusion constraints than by learning networks separately. However, the gains accomplished by sharing information through fused regression do not extend to those interactions lacking true fusion constraints, and the error remains similar to learning networks separately. Because the penalty function used in adaptive fusion resembles that of fused L2 where differences in interactions are small (Figure 3), adaptive fusion's performance on accurately fused interactions is similar to that of fused L2. Because interactions which have been unfused remain underdetermined, we are not closer to learning their values than if we learned networks separately. However, when using adaptive fusion, we are able to identify which interactions should not be similar, and therefore, which orthology mappings are not predictive of functional similarity.

@@

We obtained 1870 one-to-one orthologs from Inparanoid \cite{ostlund_inparanoid_2010}, which produced 177,650 fusion-constraints between gene interactions within the two species. This number represents only $14.7\%$ of the regulatory interaction matrix in \textit{B. subtilis} and $12.9\%$ in \textit{B. anthracis}. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.

@@

When we applied fused L2 to our subsampled \textit{B. subtilis} and \textit{B. anthracis} datasets using optimized parameters, we did not see a significant improvement over solving the networks separately. Inspection of precision recall curves shows that there is increased precision at low values of recall. Essentially, precision increases for the high confidence part of the network but not for the low confidence part of the network. As a result, the slight improvements in performance may be more useful than would be suggested by AUPR alone. Interactions suggested by genome wide analysis are useful for target prediction, and the ranking of interactions produced by a network inference algorithms is a useful guide to what order these experiments should be carried out in. The high confidence part of the network represents interactions that are likely candidates for experimental validation, so improvement in recovery of these interactions is particularly useful in guiding research.

\subsection{Within-species fusion using similarity of promoter region}
Information about the similarity of TF-gene interactions can also come from knowledge about the promoter region; this may be a better predictor of target similarity than orthology. In bacteria, genes within the same operon are under the control of the same promoter. We predicted, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We applied fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon, and showed a boost in \textit{B. subtilis} network recovery using within-species fusion (Figure 6b).

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs \cite{kundaje_integrative_2015}, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compared this approach to using fusion regression. We included, along with our \textit{B. subtilis} dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different microarray and different strain of \textit{B. subtilis} \cite{nicolas2012condition}. We compared performance when learning the networks separately, rank combining, and learning the networks simultaneously using fusion regression, and showed improvement on network inference using our fused L2 approach (Figure 6a). Importantly, our approach simultaneously learns multiple networks, rather than learning networks separately than combining into one.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We tested a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we attempted to fit gene expression as a function of transcription factor activity, which have been shown to be better predictors of gene expression \cite{arrieta-ortiz_experimentally_2015}. We estimated TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. To test the integration of this approach with our fused regression, we asssesed the integration of \texti{B. subtilis} datasets, as in figure 6a, with the incorporation of TFA estimation. We randomly divided the prior known interactions in half, and used half to learn TFA. The rest we reserved as a gold standard for validation. As in previous studies, we observed a marked improvement in network inference when using transcription factor activity, which captures activation of transcription factor through mechanisms such as dimerization and interaction with required factors, over transcription factor expression level alone \cite{fu_reconstructing_2011}. We also obtained AUPR improvement when using fused regression on TFA, and showed that our gains from sharing information across datasets using fused regression were preserved and even enhanced by using TFA.

\section{Discussion}
Gene expression data, such as microarray or RNA seq, provide information about the relationship between genes by allowing an experimenter to measure correlations in expression value over time or across conditions. Many sources of information - such as the knowledge that two genes are related through orthology or belong to the same operon - provide information about the relationships between these gene-gene relationships, without necessarily providing information about the gene-gene relationships themselves. For example, the meta information that two genes belong to the same operon suggests that they are likely to have a similar set of regulators (CITE), but does little to inform the identity of those regulators. Meta-information about the structure of gene regulatory networks - specifically which pairs of interactions are \texti{a priori} likely to be similar to one another, can provide a powerful set of constraints to improve network inference performance. We present a general framework for gene regulatory network inference that incorporates this meta-information - termed fusion constraints - and apply the technique to the problem of simultaneous inference of regulatory networks in multiple species (\textit{B. subtilis} and \textit{B. anthracis}). We develop two algorithms for solving fused network inference problems, each of which has as its core the solving of a least-squares optimization problem with novel regularization constraints.

 The first method, fused L2, penalizes the squared difference of the regulatory weights of fused interactions. These constraints improve the performance of the underdetermined network inference problem, by allowing information to be shared across sources. We apply this algorithm to the problem of network inference in two distantly related biological organisms -- \textit{B. subtilis} and \textit{B. anthracis} -- and show that network recovery is improved through the introduction of fusion constraints between pairs of orthologous genes. Existing methods for cross-species network inference operate on the conserved subset of orthologous genes [CITE]. This may be appropriate with very closely related species, but could not be applied in this domain, where a large fraction (X\% and Y\%) of each species genomes do not have orthologs. Our method, in contrast, can obtain improvements in network inference performance even when the conserved subset of genes is small. We further demonstrate the viability of fused L2 as a method for combining data from multiple experimental platforms, where fusion is between each identical regulator-gene pair. Because the algorithm we developed can accomodate constraints between arbitrary pairs of regulatory interactions, any biological prior representing information about expected regulatory similarity can be represented, even if the prior provides no information about the magnitude or direction of regulation. We demonstrate this flexibility through the novel incorporation of operon structure into the gene-regulatory network inference problem [CITE NOVELTY]. In this application, fusion reflects the assumption that genes in the same operons have similar regulators [CITE]. The ability to incorporate multiple data sets describing related processes, as well as multiple data types, in a principled manner, helps us take advantage of the breadth of experimentation in biology to better learn the structure of gene regulation. This approach is particularly interesting in light of the diversity of model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design [CITE], but without a principled mechanism for combining data from multiple sources, it is difficult to fully leverage data obtained from even a slightly different model system.

Although it is important to take advantage of the similarities of related organisms for generating improved models of gene regulation, it is also critically important to understand how systems differ from one another. Our cross-species network inference method is premised on the assumption that orthologous genes have similar regulators. This is a common assumption when comparing model systems, which is difficult to evaluate on a genome wide scale using existing methods; because network inference is typically underconstrained, fitting a network that describes a particular set of experimental observations involves sampling a single network from a large set of networks that fit the data equally (or almost equally) as well. As a result, the existence of a difference between corresponding regulatory interactions in a pair of experimentally derived networks is weak evidence that a difference truly does exist. It may be that some other pair of networks exist that fit the data nearly as well, but do not exhibit this difference. As a result, global network inference algorithms are a very weak tool for uncovering evolutionary divergence. Our method, on the other hand, explicitly favors recovering networks for which evolutionarily corresponding interactions are similar. As a result, the failure to obtain networks that confirm evolutionary conservation is strong evidence that conservation does not exist; the next best network that does exhibit conservation must fit the data much worse to have overcome the bias built into the fusion constraints.

Because our approach allows us to directly test -- and reject -- the hypothesis of conserved function, and because the enforcement of fusion constraints between non-conserved interactions may degrade network inference performance, we introduce an adaptive fusion method that decouples pairs of interactions for which expression data suggests a lack of conservation. Like the fused L2 approach, this method attempts to solve for networks which minimize a function that minimizes error in fitting the expression data plus error associated with the fusion penalty. In the case of cross-species network inference, orthology may provide a useful set of constraints on network structure. However, orthology does not always accurately predict functional similarity [CITE].When a fusion constraint encourages orthologous gene, TF pairs to share similar interactions across species, but the expression data suggests that these interaction weights should not be similar, the algorithm will arrive at some compromise between these two forces. When the disagreement arose because the genes involved have taken on new function through evolutionary divergence [CITE], network recovery would be improved had the fusion constraint not been enforced. We introduce a method -- adaptive fusion -- that attempts to learn which fusion constraints should be relaxed while the network is being learned. This method is based on minimizing a saturating penalty function on fusion constraints, similar to a class of penalties that have been developed to minimise bias in regularized regression [CITE]. The result of adaptive fusion is both a network and a new set of fusion constraints, describing the learned conservation weights (including which fusion constraints have been relaxed). For the multiple species case, relaxation of fusion constraints represents orthologs which do not share similar interactions. When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function. Adaptive fusion, therefore, is a tool for network inference as well as a method for testing network conservation.

>>>>

\subsection{\textit{B. subtilis} and \textit{B. anthracis} microarray data}
We used a dataset collected for PY79, a derivative of strain 168, available on GEO with accession number GSE67023, and a dataset using BSB1, another derivative of strain 168, available at GEO with accession number GSE27219. We used two datasets for \texti{B. anthracis}, transcription profiling during iron starvation (E-MEXP-2272 on ArrayExpress), and time series over the life cycle (E-MEXP-788 on ArayExpress).

\subsection{Orthology mapping using Inparanoid}
We ran Inparanoid to obtain orthology mapping for \textit{B. subtilis} and \textit{B. anthracis} \cite{ostlund_inparanoid_2010}

@@

\noindent The matrix $\beta$ can be solved using linear regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring multiple networks. 
\subsection{Learning transcription factor activity and using TFA as regressors}

@@

\subsection{Statistical approach}
Most network inference problems have more predictors than experimental samples ($p >> N$). Many methods have been proposed for penalized regression, such as ridge regression \cite{hoerl_ridge_1970}, and applied to genomic data \cite{waldron_optimized_2011, li_network-constrained_2008}. In our application, we consider the case where related outputs (genes) are likely to be affected by common inputs (transcription factors). Approaches using fusion combine statistical strength across the multiple regression tasks by introducing regularization functions reflecting prior knowledge on output structure representing how outputs are related \cite{kim_tree-guided_2012,Land1997,Chen2010,Petry2011,Hebiri2011}, and approaches which attempt to simultaneously learn the regression coefficients and the output structure \cite{sohn_joint_2012}. Fusion has been proposed as a method for utilizing correlation information from data by penalizing pairwise differences of coefficients \cite{Daye2009,Price2014}. Price et al. and Bilgrau et al. use a fused ridge estimator for jointly estimating multiple inverse covariance matrices \cite{Price2014a,Bilgrau2015}; we take a similar approach, adding an fused L2 penalty to our ridge regression to incorporate prior knowledge about relationships between input-output pairs. We also apply a penalty function bounded by a constant to produce unbiased estimators for large coefficients, combined with an L2 penalty, similar to SCAD-L2 \cite{Zeng2012}. This approach allows us to simultaneously learn the regression coefficients and evaluate the validity of this prior information.

@@

In order to avoid this difficulty, we use depth-first search to identify linked columns of each TF expression matrix $Y^{S_j}$, then forming design and response matrices through vectorization. We can then incorporate fusion constraints as in the case of single-source single response-variable fused regression. In most cases, we have found the direct solve using augmented matrices to be adequate. This is because the structure of orthology tends to be sparse, so only a small number of genes must be solved at once. However, in the general case of $c_i$ response variables from each of $i \in \text{species}$ data sources the design matrix to be solved is of size $\displaystyle \sum_{c_i} rows(X^{S_i}) \times \displaystyle \sum_{c_i} cols(X^{S_i})$ plus a relatively small number of rows representing fusion constraints. Because the scaling of this algorithm has a complicated depence on the constraint structure used, a general description of its runtime is difficult. However, in the case of multi-species network inference with one-to-one orthology, the network associated with each pair of orthologous genes requires solving a linear system with approximately twice as many observations and unknowns as the single species case. Linear systems of this size can be solved quickly using standard techniques. When the number of genes linked by fusion constraints becomes large, however, the augmented design matrix approach becomes impractical.

@@

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthologs are not functionally analogous  (Figure 4). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``incorrect'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``incorrect'' mappings. We verified that adaptive-fusion unfused mostly ``incorrectly fused'' interactions, while leaving truly analogous interactions fused (Figure 4a. Note that while pairs of interactions between true orthologs were sampled randomly so that the difference in interactions tended to be small, and pairs of interactions between incorrect orthologs were sampled randomly so that the difference in interactions tended to be large, there was an element of randomness in the sampling. This resulted in a small fraction of interactions between incorrect othologs being similar, and similarly, a small fraction of interactions between correct orthologs being different. Consequently, some interactions between incorrect orthologs remained fused and some interactions between correct orthologs were unfused.

We compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information (Figure 4b). Because fused L2 heavily penalizes large differences between weights which are predicted to be similar, it is able to retrieve a more accurate network for those interactions with true fusion constraints than by learning networks separately. However, the gains accomplished by sharing information through fused regression do not extend to those interactions lacking true fusion constraints, and the error remains similar to learning networks separately. Because the penalty function used in adaptive fusion resembles that of fused L2 where differences in interactions are small (Figure 3), adaptive fusion's performance on accurately fused interactions is similar to that of fused L2. Because interactions which have been unfused remain underdetermined, we are not much closer to learning their values than if we learned networks separately. However, when using adaptive fusion, we are able to identify which interactions should not be similar, and therefore, which orthology mappings are not predictive of functional similarity.

@@

We obtained 1870 one-to-one orthologs from Inparanoid \cite{ostlund_inparanoid_2010}, 95 of which are transcription factors, which produced 177,650 fusion-constraints between gene interactions within the two species. This number represents only $14.7\%$ of the regulatory interaction matrix in \textit{B. subtilis} and $12.9\%$ in \textit{B. anthracis}. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.

@@

When we applied fused L2 to our subsampled \textit{B. subtilis} and \textit{B. anthracis} datasets using optimized parameters, we did not see a significant improvement over solving the networks separately. Inspection of precision recall curves showed that there is increased precision at low values of recall. Essentially, precision increases for the high confidence part of the network but not for the low confidence part of the network. As a result, the improvements in performance may be more useful than would be suggested by AUPR alone. Interactions suggested by genome wide analysis are useful for target prediction, and the ranking of interactions produced by a network inference algorithms is a useful guide to what order these experiments should be carried out in. The high confidence part of the network represents interactions that are likely candidates for experimental validation, so improvement in recovery of these interactions is particularly useful in guiding research.

\subsection{Within-species fusion using similarity of promoter region}
Information about the similarity of TF-gene interactions can also come from knowledge about the promoter region; this may be a better predictor of target similarity than orthology. In bacteria, genes within the same operon are under the control of the same promoter \cite{lawrence_shared_2002}. We predicted, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We applied fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon, and showed a boost in \textit{B. subtilis} network recovery using within-species fusion (Figure 6b).

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs \cite{paten_nih_2015,kundaje_integrative_2015}, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compared this approach to using fusion regression. We included, along with our \textit{B. subtilis} dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different tiling microarray (vs custom microarray) and different strain of \textit{B. subtilis} \cite{nicolas2012condition}. We compared performance when learning the networks separately, rank combining, and learning the networks simultaneously using fusion regression, and showed improvement on network inference using our fused L2 approach (Figure 6a). Importantly, our approach simultaneously learns multiple networks, rather than learning networks separately than combining into one.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We tested a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we fit gene expression as a function of transcription factor activity, as shown in \textit{B. subtilis} by Arrieta-Ortiz et al \cite{arrieta-ortiz_experimentally_2015}. We estimated TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. To test the integration of this approach with our fused regression, we asssesed the combination of \texti{B. subtilis} datasets, as in figure 6a, with the incorporation of TFA estimation. We randomly divided the prior known interactions in half, and used half to learn TFA. The rest we reserved as a gold standard for validation. As in previous studies, we observed a marked improvement in network inference when using transcription factor activity, which captures activation of transcription factor through mechanisms such as dimerization and interaction with required factors, over transcription factor expression level alone \cite{fu_reconstructing_2011}. We also obtained AUPR improvement when using fused regression on TFA, and showed that our gains from sharing information across datasets using fused regression were preserved and even enhanced by using TFA.

\section{Discussion}
Gene expression data, such as microarray or RNA seq, provide information about the relationship between genes by allowing an experimenter to measure correlations in expression value over time or across conditions. Many sources of information - such as the knowledge that two genes are related through orthology or belong to the same operon - provide information about the relationships between these gene-gene relationships, without necessarily providing information about the gene-gene relationships themselves. For example, the meta information that two genes belong to the same operon suggests that they are likely to have a similar set of regulators \cite{lawrence_shared_2002}, but does little to inform the identity of those regulators. Meta-information about the structure of gene regulatory networks - specifically which pairs of interactions are \texti{a priori} likely to be similar to one another, can provide a powerful set of constraints to improve network inference performance. We present a general framework for gene regulatory network inference that incorporates this meta-information - termed fusion constraints - and apply the technique to the problem of simultaneous inference of regulatory networks in multiple species (\textit{B. subtilis} and \textit{B. anthracis}). We develop two algorithms for solving fused network inference problems, each of which has as its core the solving of a least-squares optimization problem with novel regularization constraints.

 The first method, fused L2, penalizes the squared difference of the regulatory weights of fused interactions. These constraints improve the performance of the underdetermined network inference problem, by allowing information to be shared across sources. We apply this algorithm to the problem of network inference in two distantly related biological organisms -- \textit{B. subtilis} and \textit{B. anthracis} -- and show that network recovery is improved through the introduction of fusion constraints between pairs of orthologous genes. Existing methods for cross-species network inference operate on the conserved subset of orthologous genes \cite{dillman_comparative_2015}. This may be appropriate with very closely related species, but could not be applied in this domain, where a large fraction (X\% and Y\%) of each species genomes do not have orthologs. Our method, in contrast, can obtain improvements in network inference performance even when the conserved subset of genes is small. We further demonstrate the viability of fused L2 as a method for combining data from multiple experimental platforms, where fusion is between each identical regulator-gene pair. Because the algorithm we developed can accomodate constraints between arbitrary pairs of regulatory interactions, any biological prior representing information about expected regulatory similarity can be represented, even if the prior provides no information about the magnitude or direction of regulation. We demonstrate this flexibility through the novel incorporation of operon structure into the gene-regulatory network inference problem. In this application, fusion reflects the assumption that genes in the same operons have similar regulators \cite{lawrence_shared_2002}. The ability to incorporate multiple data sets describing related processes, as well as multiple data types, in a principled manner, helps us take advantage of the breadth of experimentation in biology to better learn the structure of gene regulation. We illustrate this by combining two different \textit{B. subtilis} datasets and show that fused L2 is an improvement over current approaches to combining data \cite{marbach_revealing_2010} because of our ability to exploit the statistical power that our expanded datasets afford us. This approach is particularly interesting in light of the diversity of model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design \cite{stolfi_genetic_2012}, but without a principled mechanism for combining data from multiple sources, it is difficult to fully leverage data obtained from even a slightly different model system.

Although it is important to take advantage of the similarities of related organisms for generating improved models of gene regulation, it is also critically important to understand how systems differ from one another. Our cross-species network inference method is premised on the assumption that orthologous genes have similar regulators. Existing approaches to the genome-wide testing of this assumption learn regulatory networks separately, then compare to identify conservation \cite{aytes_cross-species_2014, Wang2014a}. But because network inference is typically underconstrained, fitting a network that describes a particular set of experimental observations involves sampling a single network from a large set of networks that fit the data equally (or almost equally) as well. As a result, the existence of a difference between corresponding regulatory interactions in a pair of experimentally derived networks is weak evidence that a difference truly does exist. It may be that some other pair of networks exist that fit the data nearly as well, but do not exhibit this difference. As a result, global network inference algorithms are a very weak tool for uncovering evolutionary divergence. Our method, on the other hand, explicitly favors recovering networks for which evolutionarily corresponding interactions are similar. As a result, the failure to obtain networks that confirm evolutionary conservation is strong evidence that conservation does not exist; the next best network that does exhibit conservation must fit the data much worse to have overcome the bias built into the fusion constraints.

Because our approach allows us to directly test -- and reject -- the hypothesis of conserved function, and because the enforcement of fusion constraints between non-conserved interactions may degrade network inference performance, we introduce an adaptive fusion method that decouples pairs of interactions for which expression data suggests a lack of conservation. Like the fused L2 approach, this method attempts to solve for networks which minimize a function that minimizes error in fitting the expression data plus error associated with the fusion penalty. In the case of cross-species network inference, orthology may provide a useful set of constraints on network structure. However, orthology does not always accurately predict functional similarity \cite{gabaldon_functional_2013}.When a fusion constraint encourages orthologous gene, TF pairs to share similar interactions across species, but the expression data suggests that these interaction weights should not be similar, the algorithm will arrive at some compromise between these two forces. If the disagreement arose because the genes involved have taken on new function through evolutionary divergence \cite{kellis_proof_2004}, network recovery would be improved had the fusion constraint not been enforced. We introduce a method -- adaptive fusion -- that attempts to learn which fusion constraints should be relaxed while the network is being learned. This method is based on minimizing a saturating penalty function on fusion constraints, similar to a class of penalties that have been developed to minimise bias in regularized regression \cite{fan2001variable,zhang2010nearly}. The result of adaptive fusion is both a network and a new set of fusion constraints, describing the learned conservation weights (including which fusion constraints have been relaxed). For the multiple species case, relaxation of fusion constraints represents orthologs which do not share similar interactions. When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function. Adaptive fusion, therefore, is a tool for network inference as well as a method for testing network conservation.

Because our model shares its basic assumptions about the role of transcription factors in gene expression dynamics with models developed for single-species network inference, we are able to leverage techniques developed for the single-species estimation of transctiption factor activity \cite{fu_reconstructing_2011}. The performance gains of this additional step in the cross-species case are significant.
